---
title: "Assessing Clinical Risk Stratification in TB Patients"
subtitle: "Can Medical Officers' Risk Assessment Be Improved by an Algorithmic Approach?"
author:
  - "Nidhi"
  - "TB Capstone Research Project"
date: today
format:
  docx:
    reference-doc: reference-doc.docx
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 2
    number-sections: false
execute:
  warning: false
  message: false
---

```{=openxml}
<w:p><w:r><w:br w:type="page"/></w:r></w:p>
```

```{r setup}
#| include: false
library(readxl)
library(dplyr)
library(tibble)
library(janitor)
library(ggplot2)
library(corrplot)

data <- read_xlsx("../data/clinical parameters_TBpatients.xlsx")
data <- clean_names(data)
```

# Background

In this project, we examine clinical data from TB patients to test whether medical
officers appropriately assess risk based on common danger signs and symptoms, or
whether an algorithmic (rule-based) approach to risk stratification would perform
better — particularly for identifying seriously ill patients who need to be
admitted or referred urgently. The analysis below is organized question by
question, following the study brief.

# Question 1: Bring the Date of Assessment Column First and Rename it to "start"

**Step 1 — Rename and relocate the column.**

```{r q1}
data <- data %>%
  rename(start = date_of_assessment) %>%
  relocate(start, .before = end)

head(data)
```

**Result:** The `date_of_assessment` column is renamed to `start` and moved to the
first position in the dataset, ahead of the `end` column.

# Question 2: Create a Unique ID from the Row ID

**Step 1 — Generate the ID column.**

```{r q2}
data <- rownames_to_column(data, var = "id")
head(data$id)
```

**Result:** Each patient record now has a unique `id` derived from its row number.

# Question 3: Mean Delay in Assessment and Mean Age

**Step 1 — Calculate assessment delay.**

```{r q3a}
data <- data %>%
  mutate(assessment_delay = as.numeric(difftime(end, start, units = "days")))
```

**Step 2 — Calculate the means.**

```{r q3b}
mean_delay <- mean(data$assessment_delay, na.rm = TRUE)
mean_age   <- mean(data$age, na.rm = TRUE)

mean_delay
mean_age
```

**Step 3 — Check for missing values.**

```{r q3c}
sum(is.na(data$assessment_delay))
sum(is.na(data$age))
```

**Step 4 — Replace missing age values with the mean.**

```{r q3d}
data <- data %>%
  mutate(age = ifelse(is.na(age), mean_age, age))
```

**Result:** The mean delay in assessment was `r round(mean_delay, 2)` days, and the
mean age of patients was `r round(mean_age, 2)` years. Missing age values were
replaced with the mean age.

# Question 4: Percentage of Patients with Cough as a Symptom

**Step 1 — Calculate the percentage.**

```{r q4}
cough_percentage <- mean(data$current_symptoms_cough) * 100
round(cough_percentage, 2)
```

**Result:** `r round(cough_percentage, 2)`% of patients reported cough as a symptom.

# Question 5: Stacked Column Chart of Patients by District and Gender

**Step 1 — Build the chart.**

```{r q5}
#| fig-width: 8
#| fig-height: 5
ggplot(data, aes(x = district, fill = gender)) +
  geom_bar() +
  labs(
    title = "Number of TB Patients by District and Gender",
    x = "District",
    y = "Number of Patients",
    fill = "Gender"
  ) +
  theme_minimal()
```

# Question 6: Does Age Follow a Normal Distribution?

**Step 1 — Visualize the distribution.**

```{r q6a}
#| fig-width: 7
#| fig-height: 5
ggplot(data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Age",
    x = "Age",
    y = "Frequency"
  ) +
  theme_minimal()
```

**Step 2 — Confirm statistically with the Shapiro-Wilk test.**

```{r q6b}
shapiro.test(data$age)
```

**Interpretation:**

- The age distribution is not symmetric.
- Most patients fall between 15 and 35 years.
- There is a long right tail extending to older ages (50–80 years), indicating a
  **positively (right) skewed** distribution rather than a bell-shaped one.
- The Shapiro-Wilk test confirms this statistically (W = 0.946, p < 0.001).

**Conclusion:** Age does **not** follow a normal distribution.

# Question 7: Stratify Patients by Number of Abnormal Parameters

Patients are stratified into risk categories based on the number of abnormal
clinical parameters:

- **1 abnormal parameter** → Low Risk
- **2–3 abnormal parameters** → Moderate Risk
- **≥4 abnormal parameters** → High Risk

**Step 1 — Create the risk category variable.**

```{r q7}
data <- data %>%
  mutate(
    risk_category = case_when(
      no_of_abnormal_parameters == 1 ~ "Low Risk",
      no_of_abnormal_parameters >= 2 & no_of_abnormal_parameters <= 3 ~ "Moderate Risk",
      no_of_abnormal_parameters >= 4 ~ "High Risk",
      TRUE ~ NA_character_
    )
  )

table(data$risk_category)
```

# Question 8: Cross-Tabulate Medical Officer Risk vs. Algorithmic Risk

This is the central question of the project: how often does a medical officer
under-classify a patient as low risk when the objective parameter count places
them at high risk?

**Step 1 — Clean up the medical officer's risk labels.**

```{r q8a}
data$mo_risk <- recode(
  data$final_risk_stratification_by_the_medical_officer,
  "Low risk- Marked for provisional follow-up later." = "Low Risk",
  "Moderate Risk- Marked for mandatory follow-up later at CHC/PHC/HWC" = "Moderate Risk",
  "High risk- Referring to District Hospital/Sub-district hospital or tertiary care facility with availability of intensive or emergency or inpatient care." = "High Risk"
)
```

**Step 2 — Cross-tabulate.**

```{r q8b}
cross_tab <- table(data$mo_risk, data$risk_category)
cross_tab
```

**Interpretation:** The cross-tabulation above shows the number of patients whom
the medical officer classified as low risk, cross-referenced against their
algorithmic (parameter-count-based) risk category. Any patients falling in the
"Low Risk" row but "High Risk" column represent cases where the medical officer
under-assessed a seriously ill patient — the key safety concern this project set
out to investigate.

# Question 9: Descending Order Chart of Abnormal Parameters

**Step 1 — Count patients with each abnormal parameter.**

```{r q9a}
# Each of these columns is a 0/1 flag per the data dictionary (1 = abnormal),
# so column sums give the number of patients with that parameter abnormal.
abnormal_counts <- data.frame(
  Parameter = c("Respiratory Rate", "BMI", "Haemoglobin", "Oxygen Saturation"),
  Count = c(
    sum(data$respiratory_rate, na.rm = TRUE),
    sum(data$bmi, na.rm = TRUE),
    sum(data$haemoglobin, na.rm = TRUE),
    sum(data$oxygen_saturation, na.rm = TRUE)
  )
) %>%
  arrange(desc(Count))

abnormal_counts
```

**Step 2 — Plot in descending order.**

```{r q9b}
#| fig-width: 7
#| fig-height: 5
ggplot(abnormal_counts,
       aes(x = reorder(Parameter, -Count), y = Count, fill = Parameter)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = Count), vjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("#0F6E6E", "#2A9D8F", "#76C7C0", "#B2DFDB")) +
  labs(
    title = "Number of Persons with Abnormal Parameters",
    x = "Clinical Parameter",
    y = "Number of Persons"
  ) +
  expand_limits(y = 380) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    axis.title = element_text(face = "bold", size = 9),
    axis.text = element_text(face = "bold", size = 7),
    legend.position = "none"
  )
```

**Result:** BMI was the most frequently abnormal parameter (352 patients),
followed by respiratory rate (171), haemoglobin (138), and oxygen saturation (57).

# Question 10: Subset of Male Patients Without Blood in Sputum

**Step 1 — Create the subset.**

```{r q10}
male_no_blood <- subset(data, gender == "Male" & blood_in_sputum == 0)
nrow(male_no_blood)
head(male_no_blood)
```

# Question 11: Additional Analysis (Optional)

*Exploratory analysis beyond the core questions.*

## Step 1: Risk Category by District

```{r q11a}
#| fig-width: 8
#| fig-height: 5
table(data$district, data$risk_category)

ggplot(data, aes(x = district, fill = risk_category)) +
  geom_bar() +
  scale_fill_manual(values = c("#0F6E6E", "#76C7C0", "#B2DFDB")) +
  labs(
    title = "Risk Category by District",
    x = "District",
    y = "Number of Patients",
    fill = "Risk Category"
  ) +
  theme_minimal()
```

## Step 2: Assessment Delay by District

```{r q11b}
#| fig-width: 7
#| fig-height: 5
delay_by_district <- aggregate(
  assessment_delay ~ district,
  data = data,
  mean,
  na.rm = TRUE
)
delay_by_district$assessment_delay <- round(delay_by_district$assessment_delay, 2)
delay_by_district

ggplot(delay_by_district,
       aes(x = reorder(district, assessment_delay), y = assessment_delay, fill = district)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("#0F6E6E", "#2A9D8F", "#76C7C0", "#B2DFDB")) +
  geom_text(aes(label = assessment_delay), vjust = -0.4, fontface = "bold", size = 4) +
  labs(
    title = "Average Assessment Delay by District",
    x = "District",
    y = "Average Delay (Days)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```

## Step 3: Distribution of Risk Categories

```{r q11c}
#| fig-width: 7
#| fig-height: 5
ggplot(data, aes(risk_category, fill = risk_category)) +
  geom_bar() +
  scale_fill_manual(values = c("#B2DFDB", "#76C7C0", "#0F6E6E")) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.4, fontface = "bold") +
  labs(
    title = "Distribution of Risk Categories",
    x = "Risk Category",
    y = "Number of Patients"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```

## Step 4: Age Distribution by Risk Category

```{r q11d}
#| fig-width: 7
#| fig-height: 5
ggplot(data, aes(risk_category, age, fill = risk_category)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#B2DFDB", "#76C7C0", "#0F6E6E")) +
  labs(
    title = "Age Distribution across Risk Categories",
    x = "Risk Category",
    y = "Age"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```

**Interpretation:**

- Moderate Risk patients had the highest median age among all risk categories.
- High Risk and Moderate Risk groups showed greater variability in age compared
  with the Low Risk group.
- A few older outliers were observed in the Low Risk and NA categories.
- The considerable overlap in age distributions suggests that age alone does not
  clearly differentiate patients across risk categories.

## Step 5: Correlation Heatmap of Clinical Parameters

```{r q11e}
#| fig-width: 6
#| fig-height: 6
vars <- data[, c("age", "bmi", "respiratory_rate", "oxygen_saturation", "haemoglobin")]
corrplot(cor(vars, use = "complete.obs"), method = "color")
```

**Interpretation:**

- Age shows a very weak positive correlation with BMI.
- Age has almost no correlation with respiratory rate, oxygen saturation, or haemoglobin.
- BMI also shows very weak correlations with the remaining variables.
- Respiratory rate, oxygen saturation, and haemoglobin do not show strong
  relationships with one another.
- Overall, none of the variables show a strong correlation (no dark blue or dark
  red cells outside the diagonal).

# Summary and Conclusion

Overall, this analysis suggests that age skews right rather than following a
normal distribution, BMI is the most commonly abnormal clinical parameter in this
population, and — most importantly for the study's core question — the
cross-tabulation in Question 8 highlights specific instances where medical
officers' subjective risk classification diverged from the objective,
parameter-count-based risk category. This supports further exploration of an
algorithmic approach as a potential safety net for identifying seriously ill TB
patients who might otherwise be under-triaged.
